---
title: "SALUS validation figures"
author: "Jill Deines"
date: "July 10, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---

some change

Goal: Make Figures for validating SALUS output for chosen models

* yield validation for 0018 model run 
* Water use for BAU and LEMA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path='../figure/01.00_salusAnalyses_validFigures/',
                      cache = FALSE)
```

**R Packages Needed**

```{r packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(lemon) # easy ggplot panel ticks
library(here)

# library(devtools)
# install_github('jdeines/salustools')
library(salustools) # for nqb theme

sessionInfo()
```


*Directories*

```{r directories}
baseDir <- here::here()

# nass yield summary
nassDir <- paste0(baseDir, '/data/tabular/NASS')

# salus output
salusDir <- paste0(baseDir, '/data/SALUS_output')
yieldValidModel <- '1_yieldCalibration_0018'
BAUmodel <- '2_BAU_0054'
LEMAmodel <- '3_LEMA_0086'

# wimas well data
wellDir <- paste0(baseDir, '/data/tabular/wellData')
wellFileName<- 'WIMAS_1996-2016_SheridanNull9_convertedUnits.csv'
```


# SALUS yield valiadation
Plot salus yield IQR against NASS, for rainfed and irrigation crops

## load and process NASS

```{r loadYieldDetailed}
plotYear <- 2008

# state of kansas
nassState <- read_csv(paste0(nassDir,
                    '/NASS_yields_STATE_KS_2006-2017_4crops_CLEANED.csv')) %>%
  mutate(type = recode(type, "NON-IRRIGATED" = "Rainfed", 
                             "IRRIGATED" = "Irrigated", 
                             "ALL" = "Combined")) %>%
  # drop unneeded "coombineds"
  filter(!(type == "Combined" & crop %in% c('CORN'))) %>%
  filter(year >= plotYear) %>%
  dplyr::select(-c(yield_bu_acre, state_name)) %>%
  rename(stateNassYields = yield_kg_ha)

# sheridan county level
nassSheridan <- read_csv(paste0(nassDir,
                    '/NASS_yields_KS_2006-2017_4crops_CLEANED.csv')) %>%
  filter(county_name == 'SHERIDAN') %>%
  mutate(type = recode(type, "NON-IRRIGATED" = "Rainfed", 
                             "IRRIGATED" = "Irrigated", 
                             "ALL" = "Combined")) %>%
  rename(sheridanNassYields = yield_kg_ha) %>%
  # drop unneeded "coombineds"
  filter(!(type == "Combined" & crop %in% c('CORN'))) %>%
  filter(year >= plotYear) %>%
  dplyr::select(c(year, crop, type, sheridanNassYields))

```

## load SALUS runs
load salus runs for chosen 0018, best yield fit with N off and water solid (75%, 25 mm)

```{r salusAllYieldsJoin}
# converter for salus output
statusKey <- data.frame(type = c('Irrigated', 'Rainfed'),
                        irrigated = c('Y','N'),
                        stringsAsFactors = FALSE)

cropKey <- data.frame(crop = c('CORN','SORGHUM','WHEAT','SOYBEANS','ALFALFA'),
                      cropsalus = c('MZ','SG','WH','SB','AL'),
                      stringsAsFactors = FALSE)

# load yield tabular results
yields <- read_csv(paste0(salusDir,'/',yieldValidModel,'/results/yields_statsAndTotals.csv')) %>%
    rename(cropsalus = crop) %>%
  left_join(cropKey, by = 'cropsalus') %>%
  left_join(statusKey, by = 'irrigated') %>%
  dplyr::select(c(year, crop, type, yield_Q1_kgha, yield_median_kgha, 
                  yield_Q3_kgha))  %>%
  filter(crop != 'ALFALFA') %>%
  filter(year >= plotYear)

# combine data
yieldValid <- yields %>%
  full_join(nassSheridan, by = c('year','crop','type')) %>%
  full_join(nassState, by = c('year','crop','type'))

# state ---------
yieldValidSimple <- yieldValid %>%
  # assign 'combined' nass data to rainfed
    mutate(type2 = recode(type, 'Combined' = 'Rainfed')) %>%
  mutate(crop = recode(crop, 'CORN' = 'Corn', 'SORGHUM' = 'Sorghum',
                       'SOYBEANS' = 'Soybeans','WHEAT' = 'Wheat'))

# extract and smoosh back together
nasser <- yieldValidSimple %>% 
  select(c(year, crop, type2, stateNassYields)) %>%
  filter(!is.na(stateNassYields))

actuallySimple <- yieldValidSimple %>%
  select(c(year, crop, type2, yield_median_kgha)) %>%
  filter(!is.na(yield_median_kgha)) %>%
  left_join(nasser, by = c('year','crop','type2'))

# county ------------
# assoicate nass state combined with salus rainfed
yieldValidSimple2 <- yieldValid %>%
  # assign 'combined' nass data to rainfed
    filter(type != 'Combined') %>%
  rename(type2 = type) %>%
  mutate(crop = recode(crop, 'CORN' = 'Corn', 'SORGHUM' = 'Sorghum',
                       'SOYBEANS' = 'Soybeans','WHEAT' = 'Wheat'))
# extract and smoosh back together
nasserCounty <- yieldValidSimple2 %>% 
  select(c(year, crop, type2, sheridanNassYields)) %>%
  filter(!is.na(sheridanNassYields))

# assoicate nass state combined with salus rainfed
yieldValidState <- yieldValidSimple %>%
  select(c(year, crop, type2, yield_median_kgha)) %>%
  filter(!is.na(yield_median_kgha)) %>%
  left_join(nasser, by = c('year','crop','type2'))

actuallySimpleCounty <- yieldValidSimple %>%
  select(c(year, crop, type2, type, yield_median_kgha)) %>%
  filter(type != 'Combined') %>%
  left_join(nasserCounty, by = c('year','crop','type2')) %>%
    filter(!is.na(sheridanNassYields)) 
```

## Plot!
yield validation

```{r yieldValidSimple_manuscript, dev=c('png'), fig.height = 3.5, fig.width =3.5, dpi=600}
# plot it
# plot
ggplot(actuallySimple,
       aes(x=stateNassYields/1000, y = yield_median_kgha/1000, color=crop)) +
      stat_smooth(method = 'lm', se=F, size=1, colour='cornflowerblue') +
  geom_point(aes(shape = type2)) +
  # county data - outline with black
      geom_point(aes(x=sheridanNassYields/1000, y = yield_median_kgha/1000, shape = type2), color = 'black', size = 2.5, data = actuallySimpleCounty) +
  geom_point(aes(x=sheridanNassYields/1000, y = yield_median_kgha/1000, shape = type2, color = crop), data = actuallySimpleCounty) +
  scale_shape_manual(values = c(16, 17)) +
    geom_abline(slope = 1, intercept = 0, linetype='dashed') + 
     coord_equal(xlim=c(0,15), ylim=c(0,15)) + 
  scale_color_manual(values=c('goldenrod2','darkorange2','forestgreen','orange4'))+
  ylab('SALUS Yields (t/ha)') + xlab('NASS Yields (t/ha)') +
   theme_bw() + theme(legend.position = c(.18,.69),
                      legend.background = element_blank(),
                     legend.title = element_blank(),
                     legend.key = element_blank(),
                    panel.grid = element_blank(),
                     axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))


# and get rsq - state
simpleAgreement <- lm(yield_median_kgha~stateNassYields, data=actuallySimple)
summary(simpleAgreement)

# rsq with county
simpleAgreement <- lm(yield_median_kgha~sheridanNassYields, data=actuallySimpleCounty)
summary(simpleAgreement)
```

## Quantify SALUS and NASS yield agreement

```{r yieldNums}
precDiffs <- yieldValid %>%
  select(-c(yield_Q1_kgha,yield_Q3_kgha)) %>%
  # calculate percent difference
  mutate(percDiff = ((yield_median_kgha - stateNassYields) / stateNassYields)*100) %>%
  arrange(crop,type,year)

precDiffs %>%
  filter(type == 'Irrigated') %>%
  group_by(crop) %>%
  summarize(maxDisagree = max(abs(percDiff), na.rm=T))
```


# Water validation figure

## load and format data

### WIMAS well data

```{r loadWimas}
# load wimas data for sheridan 6
wimas <- read_csv(paste0(wellDir, '/', wellFileName)) %>%
  filter(masterid == 'sheridan')  

# count wells in study area duing study period
wimas %>% 
  filter(year > 2005 & year < 2018) %>%
  filter(!is.na(volume_m3)) %>%
  group_by(PDIV_ID) %>%
  slice(1) %>%
  nrow()

# get annual volume totals and format to join with SALUS
wimasAnnual <- wimas %>%
  group_by(year) %>%
  summarise(wimasVolume_m3 = sum(volume_m3, na.rm=TRUE),
            wimasArea_m2 = sum(area_m2, na.rm=TRUE),
            wimasDepth_mm = mean(depth_m * 1000, na.rm=TRUE)) %>%
  mutate(wimasArea_ha = wimasArea_m2 * 1e-4,
         wimasVolume_km3 = wimasVolume_m3 * 1e-9,
         datasource = 'WIMAS') %>%
  dplyr::select(c(year, wimasVolume_km3))
```

### SALUS water results
load and combine with WIMAS

```{r loadSALUS}
bau <- read_csv(paste0(salusDir,'/', BAUmodel, '/results/WaterVars_meansTotals.csv')) %>%
  mutate(runID = '0054')
lema <- read_csv(paste0(salusDir,'/', LEMAmodel, '/results/WaterVars_meansTotals.csv')) %>%
  mutate(runID = '0086')

water <- bau %>%
  bind_rows(lema) %>%
  mutate(datasource = 'salus') %>%
  filter(variable == 'irrigation') %>%
  mutate(salusVolume_km3 = totalVolume_m3 * 1e-9) %>%
  select(c(runID, year, salusVolume_km3))

master <- water %>%
  left_join(wimasAnnual, by = c('year'))

# assume a 90% efficiency -------------------------------
master$salusVol_90adj <- master$salusVolume_km3 / 0.9

plotYear <- 2008
summary <- master %>%
  filter(year >= plotYear & runID %in% c('0054','0086')) %>%
  mutate(runID = recode(runID, '0086' = 'LEMA', '0054' = 'BAU'))
```



## Plot Pumping: WIMAS and Runs

```{r waterValidComparison, fig.width = 6.5, fig.height = 2.8, dev = c('png')}

# water: remove LEMA scenario pre-lema data
waterplot <- summary %>%
  filter(!(runID == 'LEMA' & year < 2013)) 

# convert km3 to million m3 with 1e3
ggplot(waterplot,
       aes(x=year, y = salusVol_90adj*1e3, color = runID, group = runID)) +
  geom_line(lwd=1.1) + 
  # add wimas
  geom_line(data = waterplot, 
      aes(x=year, y = wimasVolume_km3*1e3), color = 'blue', linetype='dashed') +
  geom_vline(xintercept = 2012.5, color = 'gray50', linetype = 'longdash') +
  scale_x_continuous(breaks = c(2008,2010,2012,2014,2016)) +
   scale_color_manual(values=rev(c("#35978f", "#bf812d"))) +
  ylab(expression(paste('Pumping Volume (million ',m^3,')',sep=''))) +
  xlab('') +
  theme_bw() + theme(legend.position = c(.2,.2), 
                     legend.title = element_blank(),
                    panel.grid = element_blank(),
                     axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))
```

## Estimated water savings
How much less pumping occurred in the BAU model vs the LEMA model?

### aggregate 5 year total

```{r modeledWaterSavings}
waterplot %>%
  filter(year >= 2013)

diffInTotalWater <- waterplot %>%
  filter(year >= 2013) %>%
  group_by(runID) %>%
  summarize(totalWaterSALUS = sum(salusVol_90adj))
diffInTotalWater

# magnitude change
diffInTotalWater[diffInTotalWater$runID == 'BAU',2] - 
  diffInTotalWater[diffInTotalWater$runID == 'LEMA',2]

# percent change
(1-(diffInTotalWater[diffInTotalWater$runID == 'LEMA',2]/diffInTotalWater[diffInTotalWater$runID == 'BAU',2])) * 100

```

### mean annual

```{r meanAnnualReductions}
# mean annual lema pumpnig
diffInTotalWater[diffInTotalWater$runID == 'LEMA',2] / 5

# mean annual BAU pumpnig
diffInTotalWater[diffInTotalWater$runID == 'BAU',2] / 5

# percent change
(1-((diffInTotalWater[diffInTotalWater$runID == 'LEMA',2] /5)/
      (diffInTotalWater[diffInTotalWater$runID == 'BAU',2]/5))) * 100

# absolute reduction annualized?
(diffInTotalWater[diffInTotalWater$runID == 'BAU',2] / 5 ) - (
diffInTotalWater[diffInTotalWater$runID == 'LEMA',2] / 5)
```






