---
title: 'Salus Analyses: Water Budget'
author: "Jill Deines"
date: "July 11, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---

Goal: Compare Water Budget between scenarios. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path='../figure/02.00_salusAnalyses_waterBudget/',
                      cache = FALSE)
```

**R Packages Needed**

```{r packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(lemon) # easy ggplot panel ticks
library(raster)
library(rgdal)
library(here)
library(cowplot)

# library(devtools)
# install_github('jdeines/salustools')
library(salustools)

sessionInfo()
```

*Directories*

```{r directories}
# salus results folders
baseDir <- here::here()
salusOutputDir <- paste0(baseDir, '/data/SALUS_output')
BAUmodel <- '2_BAU_0054'
LEMAmodel <- '3_LEMA_0086'
```


# load data
BAU and LEMA model runs

```{r loadData}
bau <- read_csv(paste0(salusOutputDir,'/', BAUmodel, '/results/WaterVars_meansTotals.csv')) %>%
  mutate(runID = '0054')
lema <- read_csv(paste0(salusOutputDir,'/', LEMAmodel, '/results/WaterVars_meansTotals.csv')) %>%
  mutate(runID = '0086')

water <- bau %>%
  bind_rows(lema) %>%
  # recalc totalVolume_km3 from m3
  dplyr::select(-c(totalVolume_km3)) %>%
  mutate(totalVolume_km3 = totalVolume_m3 * 1e-9) %>%
  dplyr::select(c(runID, year, variable, totalVolume_km3, depth_mean_mm)) %>%
  mutate(runID = recode(runID, '0086' = 'LEMA', '0054' = 'BAU')) %>%
  filter(year >=2013) 

# adjust irrigation volume only by 90% efficiency ---------------

# account for 90% efficiency in salus estimated water volumes
adjustedIrr <- water %>% filter(variable == 'irrigation') %>%
  mutate(totalVolume_km3 = totalVolume_km3 / .9)

# replace in data frame
waterIrrAdj <- water %>% filter(variable != 'irrigation') %>%
  bind_rows(adjustedIrr)
```

# recharge numbers

```{r rechargeChange}
rechargeVols <- waterIrrAdj %>%
  filter(variable == 'recharge') %>%
  group_by(runID) %>%
  summarize(meanAnnualRechage = mean(totalVolume_km3),
            totalRecharge = sum(totalVolume_km3))
rechargeVols

# percent decrease total
(1 - (rechargeVols[2,3] / rechargeVols[1,3] )) * 100
```


# Plot Water Budget Components

## annual water budgets - transpiration
for whole region (including rainfed)

epac = plant transpiration
esac = soil transpiration
recharge = deep percolation
rofc = runoff

```{r waterBudget_byYear, fig.width = 5, dev = c('png')}
unique(waterIrrAdj$variable)

# muliply km3 by 1e3 for legibility (million m3)
ggplot(waterIrrAdj %>% filter(!variable %in% c('etac','rofc','precip')),
       aes(x = year, y = totalVolume_km3*1e3, fill = runID)) +
  geom_bar(stat='identity', position = position_dodge()) +
  scale_fill_manual(values=rev(c("#35978f", "#bf812d"))) +
  facet_rep_wrap(~variable) +
  ylab(expression(paste('Volume (million ',m^3,')',sep=''))) +
  theme_bw() + theme(panel.grid = element_blank(),
                     legend.title = element_blank(),
                     legend.position = c(.1,.3),
                      axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))

# summarize epac and esac by year
waterIrrAdj %>% filter(variable %in% c('esac','epac','rofc')) %>%
  tidyr::spread(., key = runID, value = totalVolume_km3) %>%
  mutate(percdifference = ((LEMA - BAU)/BAU)*100,
         percDecrease = (1 - (LEMA/BAU))*100) %>%
  group_by(variable) %>%
  summarize(meanPercDiff = mean(percdifference),
            sdPercDiff = sd(percdifference),
            meanDecrease = mean(percDecrease),
            sdDecrease = sd(percDecrease))

# summarize epac and esac 5 year
waterIrrAdj %>% filter(variable %in% c('esac','epac','rofc')) %>%
  tidyr::spread(., key = runID, value = totalVolume_km3) %>%
  group_by(variable) %>%
  summarize(BAU = sum(BAU),
            LEMA = sum(LEMA)) %>%
  mutate(percDecrease = (1 - (LEMA/BAU))*100) 
  

# total change for these three
waterIrrAdj %>% filter(variable %in% c('esac','epac','rofc')) %>%
  tidyr::spread(., key = runID, value = totalVolume_km3) %>%
  #group_by(variable) %>%
  summarize(BAUtotal = sum(BAU),
            LEMAtotal = sum(LEMA)) %>%
  mutate(saved = BAUtotal-LEMAtotal)


```

note SUMMARIES WERE FOR WHOLE REGION...

## compare irrigation vs deep percolation
irrigation is the total water extracted (salus estimates adjusted by 90% efficiency) and deep percolation is the total amount calculated as drainage by salus across the region, including rainfall in both irrigated and dryland crops

deep percolation is called "recharge" in salus outputs

```{r inOut_byRegion, fig.width = 5, fig.height=2.75, dev = c('png')}
# keep only irrigation and recharge, and make irrigation negative
inOutData <- waterIrrAdj %>%
  filter(variable %in% c('irrigation','recharge'))

ggplot(inOutData,
   aes(x = year, y = totalVolume_km3 * 1e3, fill = variable)) +
  geom_bar(stat='identity', position = position_dodge()) +
  #scale_fill_manual(values=rev(c("#35978f", "#bf812d"))) +
  facet_rep_wrap(~runID) +
    ylab(expression(paste('Volume (million ',m^3,')',sep=''))) +
  theme_bw() + theme(panel.grid = element_blank(),
                     legend.title = element_blank(),
                     legend.position = c(.13,.8),
                      axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))
```

plot difference - by year

```{r netChange_byYear, fig.width = 4, fig.height = 2.25}
netDiff <- inOutData %>% dplyr::select(-c(depth_mean_mm)) %>%
  tidyr::spread(., key = variable, value = totalVolume_km3) %>%
  mutate(netChange = recharge - irrigation)

# totals
netDiff %>%
  group_by(runID) %>%
  summarize(pumping = sum(irrigation),
            recharge = sum(recharge),
            netChange = sum(netChange))

# by year
p1 <- ggplot(netDiff,
       aes(x=year, y = netChange * 1e3,fill=runID)) +
  geom_bar(stat='identity', position = position_dodge()) +
  ylab(expression(paste('Net Change (million ',m^3,')',sep=''))) +
  scale_fill_manual(values=rev(c("#35978f", "#bf812d"))) +
  geom_hline(yintercept = 0) +
  theme_bw() + theme(panel.grid = element_blank(),
                     legend.title = element_blank(),
                      axis.title.x = element_blank(),
                      axis.text.x = element_blank(),
                     legend.position = c(.13,.8),
                      axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))
p1
```

plot difference - over all

```{r netChange_overall, fig.height = 3.5, fig.width = 2, dev=c('png')}
totalNet <- netDiff %>%
  group_by(runID) %>%
  summarize(totalNet = sum(netChange))

p2 <- ggplot(totalNet,
       aes(x=runID, y = totalNet * 1e3,fill=runID)) +
  geom_bar(stat='identity', position = position_dodge()) +
  ylab(expression(paste('Cumulative 5-Year Net Change (million ',m^3,')',sep=''))) +
  xlab('') +
  scale_fill_manual(values=rev(c("#35978f", "#bf812d"))) +
  theme_bw() + theme(panel.grid = element_blank(),
                     legend.title = element_blank(),
                     legend.position = 'none',
                      axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))
p2 

# increase in aquifer recharge:
changeInRecharge <- totalNet$totalNet[2] - totalNet$totalNet[1]
changeInRecharge * 1e3 # million m3

# km3 / km2 = km * 1000 = m
totalDepthChange <- (changeInRecharge / 256) * 1000 # meters
totalDepthChange # m 
(totalDepthChange * 1000) #mm

# depth change per year in mm
(totalDepthChange * 1000) /5
```


Line plot of diference over time

```{r netChangeLine, fig.width = 4, fig.height = 1.25, dev=c('png')}
# calculate bau vs lema net over time
netDiffRegions <- netDiff %>%
  dplyr::select(-c(irrigation, recharge)) %>%
  tidyr::spread(., key = runID, value = netChange) %>%
  mutate(regionDiff = LEMA - BAU)

# by year
p3 <- ggplot(netDiffRegions,
       aes(x=year, y = regionDiff*1e3)) +
  geom_line() +
 ylab(expression(paste(LEMA[net],' - ', BAU[net],sep=''))) +
    scale_x_continuous(limits = c(2012.5,2017.5)) +
  #xlab('') +
    geom_hline(yintercept = 0, linetype = 'dotted') +
  
  theme_bw() + theme(panel.grid = element_blank(),
                     legend.title = element_blank(),
                     legend.position = c(.13,.8),
                     # axis.title.x = element_blank(),
                     # axis.text.x = element_blank(),
                      axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))
p3
```


stack line and bar

```{r netChangeCombine, fig.width = 4, fig.height = 3.5, dev=c('png')}

plot_grid(p1,p3, align='v', nrow=2, rel_heights=c(.6,.4))

```


# Plot Water Budget Components
for supplement

## annual water budgets
for whole region (including rainfed)

```{r waterBudget_byYear_all_supplement, fig.width = 6.5, dev = c('png')}
unique(water$variable)

# better names
waterKey <- data.frame(variable = c('epac','esac','irrigation','recharge','rofc','etac','precip'),
                       variableNames = c('a) Plant Transpiration', 'b) Soil Evaporation','d) Irrigation','e) Recharge','c) Runoff','ET','f) Annual Precip'))

waterIrrAdj2 <- waterIrrAdj %>%
  left_join(waterKey)

ggplot(waterIrrAdj2 %>% filter(!variable %in% c('etac')),
       aes(x = year, y =depth_mean_mm, fill = runID)) +
  geom_bar(stat='identity', position = position_dodge()) +
  scale_fill_manual(values=rev(c("#35978f", "#bf812d"))) +
  facet_rep_wrap(~variableNames) +
  ylab(expression(paste('Regional Mean (mm)',sep=''))) +
  theme_bw() + theme(panel.grid = element_blank(),
                     legend.title = element_blank(),
                     legend.position = c(.1,.3),
                      axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     axis.title=element_text(size=11))
```

