---
title: 'Salus Analyses: Economics'
author: "Jill Deines"
date: "July 11, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---

Goal: Calculate monetary differences between pre-lema irrigation behavior model and LEMA irrigation behavior model

* income: crop yield * commodity prices
* energy savings and cost


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path='../figure/04.00_salusAnalyses_money/',
                      cache = FALSE)
```

**R Packages Needed**

```{r packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)

sessionInfo()
```

*Directories*

```{r directories}
baseDir <- here::here()

# price data ------------
priceDir <- paste0(baseDir,'/data/tabular/pricing/commodityPrices')
# filename for inflation adjustment table (made in inflationTable chunk)
cpiFile <- 'cpi_priceAdjustments.csv'
# prices downloaded from NASS
nassRawFile <- 'NASS_price_commodity5_2000-2017.csv'


# salus output ------------------
salusDir <- paste0(baseDir, '/data/SALUS_output')
BAUmodel <- '2_BAU_0054'
LEMAmodel <- '3_LEMA_0086'

# wimas well data
wellDir <- paste0(baseDir, '/data/tabular/wellData')
wellFileName<- 'WIMAS_1996-2016_SheridanNull9_convertedUnits.csv'
```

# Load stuff

## price data
Price data was downloaded manually from the NASS quickstats tool (https://quickstats.nass.usda.gov/) for the following query: SURVEY - CROPS - FIELD CROPS - CORN/SOYBEANS/WHEAT/HAY/SORGHUM - PRICE RECEIVED - 

* CORN, GRAIN - PRICE RECEIVED, MEASURED IN $ / BU
* HAY, PRICE RECEIVED, MEASURED IN $ / TON (short ton)
* SORGHUM, GRAIN - PRICE RECEIVED, MEASURED IN $ / CWT (hundredweight, == 100 lbs)
* SOYBEANS - PRICE RECEIVED, MEASURED IN $ / BU
* WHEAT - PRICE RECEIVED, MEASUTED IN $ / BU

- TOTAL - NATIONAL - US TOTAL - 2000:2017 - ANNUAL - YEAR


Here, I load the data and clean it to have consistent pricing across crops and years

Weights of crops per bushel:

UNC site says 

* corn, shelled: 56 lbs per bushel  
* corn, ear: 70
* corn (in ear):35

crop | pounds/bushel | 1 bu/acre = 
--------------- | ------ | --------
corn/sorghum | 56 | 62.89 kg/ha
soy/wheat | 60 | 67.11 kg/ha

* https://www.extension.iastate.edu/agdm/wholefarm/html/c6-80.html
* http://www.unc.edu/~rowlett/units/scales/bushels.html

sorghum: CWT = 100 lbs

1 lb = 0.4536 kg
1 kg = 2.2 lbs

```{r nassPrice_marketingYearPeriod}

# make a units columns 
unitkey <- data.frame(Commodity = c('CORN','HAY','SORGHUM','SOYBEANS','WHEAT'),
                      unit = c('BU','TON','CWT','BU','BU'),
                      stringsAsFactors = FALSE)

# load consumer price index adjustments
cpi <- read_csv(paste0(priceDir, '/',cpiFile))

# load nass
price0 <- read_csv(paste0(priceDir, '/', nassRawFile)) %>%
  left_join(unitkey, by = c('Commodity')) %>%
  dplyr::select(c(Commodity, Year, Value, unit))

# add the adjustment factor
price <- merge(price0, cpi[,c('cpi_year','adj_factor_2017')], 
               by.x = 'Year', by.y = 'cpi_year')

# adjust prices (note: matched trials at https://data.bls.gov/cgi-bin/cpicalc.pl)
price$price_2017 <- price$Value / price$adj_factor_2017

# adjust to price per kg
kgConverter <- data.frame(Commodity = c('CORN','WHEAT','SOYBEANS','SORGHUM'),
                 nassToKgConversion = c(0.039,  0.0367, 0.0367, 0.0045),
                 stringsAsFactors = F)
priceKg <- price %>%
  left_join(kgConverter, by = c('Commodity')) %>%
  mutate(price_2017_kg = price_2017 * nassToKgConversion) %>%
  filter(Commodity != 'HAY')

# plot it
ggplot(data = priceKg, 
       aes(x=Year, y = price_2017, color = Commodity, group = Commodity)) +
  ylab('2017 US Dollars per Unit') +
  geom_line() +
  theme_bw() +
  ggtitle('Inflation-adjusted Prices')

ggplot(data = priceKg, 
       aes(x=Year, y = price_2017_kg, color = Commodity, group = Commodity)) +
  ylab('2017 US Dollars per Kg') +
  geom_line() +
  theme_bw() +
  ggtitle('Inflation-adjusted Prices, $ / kg')
```

## Energy Prices
https://www.npr.org/sections/money/2011/10/27/141766341/the-price-of-electricity-in-your-state - KS wide = 10.5 cents per kWh
https://www.electricitylocal.com/states/kansas/

* commericial: 9.24
* industrial: 7.09
* residential: 11.24

## SALUS data

### total region yields

```{r salusYields}
cropKey <- data.frame(cropFull = c('CORN','SORGHUM','WHEAT','SOYBEANS','ALFALFA'),
                      crop = c('MZ','SG','WH','SB','AL'),
                      stringsAsFactors = FALSE)


bau <- read_csv(paste0(salusDir,'/', BAUmodel, '/results/yields_statsAndTotals.csv')) %>%
  mutate(runID = '0054')
lema <- read_csv(paste0(salusDir,'/', LEMAmodel, '/results/yields_statsAndTotals.csv')) %>%
  mutate(runID = '0086')

yield0 <- bau %>%
  bind_rows(lema) %>%
  left_join(cropKey, by = 'crop') %>%
  filter(cropFull != 'ALFALFA') %>%
  filter(irrigated == 'Y') %>%
  mutate(runID = recode(runID, '0086' = 'LEMA', '0054' = 'BAU')) %>%
  dplyr::select(c(runID, year, cropFull, yield_total_kg)) %>%
  rename(totalYield_kg =  yield_total_kg)
```

### total water use
based on summary output from 04.10_salusAnalyses_waterSA.Rmd

```{r salusWater}
# load combined wimas/salus water data, median salus yields, nass yields
# load wimas data for sheridan 6
wimasAnnual <- read_csv(paste0(wellDir, '/', wellFileName)) %>%
  filter(masterid == 'sheridan') %>%
  group_by(year) %>%
  summarise(wimasVolume_m3 = sum(volume_m3, na.rm=TRUE),
            wimasArea_m2 = sum(area_m2, na.rm=TRUE),
            wimasDepth_mm = mean(depth_m * 1000, na.rm=TRUE)) %>%
  mutate(wimasArea_ha = wimasArea_m2 * 1e-4,
         wimasVolume_km3 = wimasVolume_m3 * 1e-9,
         datasource = 'WIMAS') %>%
  dplyr::select(c(year, wimasVolume_km3))


# salus
bau2 <- read_csv(paste0(salusDir,'/', BAUmodel, '/results/WaterVars_meansTotals.csv')) %>%
  mutate(runID = '0054')
lema2 <- read_csv(paste0(salusDir,'/', LEMAmodel, '/results/WaterVars_meansTotals.csv')) %>%
  mutate(runID = '0086')

water <- bau2 %>%
  bind_rows(lema2) %>%
  filter(year >= 2013) %>%
  mutate(datasource = 'salus') %>%
  filter(variable == 'irrigation') %>%
  mutate(runID = recode(runID, '0086' = 'LEMA', '0054' = 'BAU'),
         salusVolume_km3 = totalVolume_m3 * 1e-9,
         salusVol_90adj =  salusVolume_km3 / 0.9) %>%
  select(c(runID, year, salusVolume_km3, salusVol_90adj)) %>%
  left_join(wimasAnnual, by = c('year'))

```

# Calculate income loss and savings

```{r yieldprofit}
# crop profit
priceSlim <- priceKg %>%
  rename(crop = Commodity,
         year = Year) %>%
  dplyr::select(c(year, crop, price_2017_kg))

yieldPrices <- yield0 %>% 
  rename(crop = cropFull) %>%
  left_join(priceSlim, by = c('year','crop')) %>%
  mutate(cropProfit = totalYield_kg * price_2017_kg) %>%
  filter(year >= 2013)

ggplot(yieldPrices,
       aes(year, cropProfit, color = runID, group = runID)) +
  geom_line() +
  facet_wrap(~crop)+
  theme_bw()
yieldPrices

# total diff
cropProfit <- yieldPrices %>% 
  group_by(runID, year) %>%
  summarise(cropIncome = sum(cropProfit)) %>%
  tidyr::spread(., runID, cropIncome) 
cropProfit

# difference
cropLossByYear <- cropProfit %>% 
  mutate(cropIncomeLoss = LEMA - BAU)
cropLossByYear

# money lost, 2013-2017 period
totalMoneyLostCrop <- sum(cropLossByYear$cropIncomeLoss)
totalMoneyLostCrop

# total percent decline in income
cropLossByYear %>%
  summarize(BAU_totalincome = sum(BAU),
            LEMA_totalincome = sum(LEMA))
(1-(80491071/81095452)) * 100
```

### crop water use
* 1 kwh = 3.6 MJ

* commericial: 9.24
* industrial: 7.09
* residential: 11.24

```{r watercosts}

# water costs -----------------------------
energyDollarsPerkWh <- 7.09 * 1e-2
energyDollarsPerMJ <- energyDollarsPerkWh / 3.6

waterCosts <- water %>%
  mutate(volume_m3 = salusVol_90adj * 1e9,
         MJ = volume_m3 * 3.1) %>%
  mutate(waterCosts = MJ * energyDollarsPerMJ)
waterCosts

ggplot(waterCosts, 
       aes(year, waterCosts, color = runID, group = runID)) +
  geom_line() +
  theme_bw()

# total change in water costs
yearlySavings <- waterCosts %>%
  dplyr::select(c(runID, year, waterCosts)) %>%
  tidyr::spread(., key = runID, value = waterCosts) %>%
  mutate(savings = BAU - LEMA)
yearlySavings

# total savings
totalWaterSaved <- sum(yearlySavings$savings)
totalWaterSaved

# percent savings
yearlySavings %>%
  summarize(BAU_totalWaterCosts = sum(BAU),
            LEMA_totalWaterCosts = sum(LEMA))
(1-(7272005/9691516)) * 100
```

Net savings

```{r netSavings}
totalWaterSaved + totalMoneyLostCrop
```

